---
title: "Fig.1 code"
author: "Ngoc Nguyen"
date: "2025-10-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load libraries
library(ggplot2)
library(MASS) 
library(reshape2) 
library(ggrepel)
library(reshape)
library(reshape2)
library(tidyverse)
library(dbplyr)
library(dplyr)
library(foreach)
library(doParallel)
library(purrr)
library(broom)
library(scales)
library(gghalves)
library(ggplot2)
library(rnaturalearth)
library(sf)
library(terra)    
library(tidyr)      
library(patchwork)

#Function to scale variables to 0-1 range
scale_01 <- function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

#Function to calculate the correlation of detrended x and y variables
detrend_linear <- function(a1, a2) {
  length_array <- length(a1)
  index <- 1:length_array
  
  df_lm <- data.frame(a1, a2, index)
  df_lm <- na.omit(df_lm)
  
  lm_a1 <- lm(a1 ~ index, df_lm)
  lm_a2 <- lm(a2 ~ index, df_lm)

  a1_fit <- predict(lm_a1)
  a2_fit <- predict(lm_a2)
  a1_detrend <- a1 - a1_fit
  a2_detrend <- a2 - a2_fit

  cor_df <- cor.test(a1_detrend, a2_detrend, method = "pearson")    
  r_detrend <- cor_df$estimate
  p_detrend <- cor_df$p.value
  
  p_linear_a1 <- summary(lm_a1)$coefficients[2,4]
  p_linear_a2 <- summary(lm_a2)$coefficients[2,4]
    
  return(c(r_detrend, p_detrend, p_linear_a1, p_linear_a2))
}

#Function to calculate the correlation of x and y variables
linear <- function(df) {
  cor_df <- cor.test(df$gpp, df$cWood, method = "pearson")    
  r <- cor_df$estimate
  p <- cor_df$p.value
    
  return(c(r, p))
}

```


#Calculate correlation of detrended delta cWood-GPP from TRENDY models (temporal and spatial matched with observational references published in Cabon et al. (2022))
```{r}
#Open datasets from Cabon et al. to extract the year range of data from each site
year_info <- "/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/data/cabon/Cabonetal_site_info.csv"
df_info <- read.csv(year_info)
df_info$SITE_ID <- df_info$Site
df_info <- subset(df_info, On.site.RW == TRUE) #df_info contains Start and End date of measurements

#create a list of model name and scenario to run the loop
model_name_list <- c("CLASSIC", "CLASSIC-N", "CABLE-POP", "CLM5.0", "ISBA-CTRIP", "LPX-Bern", "ORCHIDEE", "ORCHIDEEv3")
var_name = "cWood"
scenario = "S2"

for( i in 1:length(model_name_list)) {
  #Open cWood and GPP annual yearly mean for 30 locations in each model
  path1 <- paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", model_name_list[i], "/gpp_", scenario, "_31_site_weighted_yearly_mean.csv", sep = "")
  path2 <- paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/",  model_name_list[i], "/", var_name, "_", scenario, "_31_site_weighted_yearly_mean.csv", sep = "")
  
  #Open the list of coordinates of tower-overlapping grid cells for each model
  coord_path <- paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", model_name_list[i], "/", model_name_list[i], "_site_info.csv", sep = "")
    
  df_gpp <- read.csv(path1)
  df_var <- read.csv(path2)
  df_coord <- read.csv(coord_path)
  
    if(model_name == "CLM5.0") {
    df_gpp_sub <- df_gpp %>% filter(!is.na(gpp) & year < 2020)
    df_var_sub <- df_var %>% filter(!is.na(.data[[var_name]])  & .data[[var_name]] != 0 & year < 2020)
  } else {
    #Filter for (1) no NA gpp, cWood, (2) no zero cWood
    df_gpp_sub <- df_gpp %>% filter(!is.na(gpp))
    df_var_sub <- df_var %>% filter(!is.na(.data[[var_name]])  & .data[[var_name]] != 0)
  }
  

  # Merge GPP and cWood datasets and rescaling the variables
  df_merge <- inner_join(df_gpp_sub, df_var_sub, by = c("year", "lat", "lon"))
  
  df_merge <- df_merge %>%
      arrange(lat, lon, year) %>% 
      group_by(lat, lon) %>%
      mutate(biomass_increment = .data[["cWood"]] - lag(.data[["cWood"]]),) %>%
      ungroup()
  
  df_merge <- full_join(df_merge, df_coord, by = c("lat", "lon")) #Only select the coordinates satisfied the criteria mentioned above to make sure that all pixels selected do not have influence of noises, disturbances, etc.
  df_merge <- full_join(df_merge, df_info[, c("Start", "End", "SITE_ID")], by = c("SITE_ID")) #only select the years whose measurement is available
  
  model_stats <- df_merge %>% 
    group_by(SITE_ID) %>%
    filter(year >= Start & year <= End) %>% 
    ungroup() %>%
    filter(
      !is.na(cWood),
      cWood != 0,
      !is.na(gpp)
    ) %>%
    group_by(lat, lon, SITE_ID) %>%
    summarise(
      n = n(),
      r_detrend_delta = if (n > 3) detrend_linear(biomass_increment, gpp)[1] else NA,
      p_detrend_delta = if (n > 3) detrend_linear(biomass_increment, gpp)[2] else NA,
      .groups = "drop"
    ) %>%
    filter(!is.na(r_detrend_delta)) 
  
  #create path to save the dataset
  path_save <- paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/datasets/",  model_name_list[i], "_", "detrended_r_normal_delta_31_gpp_TRENDYv10_temporalMatch_not_scale", scenario, ".csv", sep = "")
  write.csv(model_stats, path_save)
}

```


#Extract correlation obtained from Cabon et al. for 31 site locations
```{r}
#Read the ring width and gpp annual files
df_rw <- read.csv("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/data/cabon/rw_onsite.csv")
df_gpp <- read.csv("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/data/cabon/flux_onsite.csv")
#mSelect only for data with QC > 0.7 and aggregate GPP by year and site categories
df_gpp_year <- aggregate(GPP_dtrd ~ Year + Site, data = subset(df_gpp, QC > 0.7), FUN = mean)
#combine RW and GPP datasets
merge_flux_rw <- list(df_rw, df_gpp_year) %>% reduce(full_join, by = c("Year", "Site"))
#get rid of NA values
merge_flux_rw <- na.omit(merge_flux_rw)
#loop through all sites to extract the correlation and p-value between detrended RW and GPP
site_name <- unique(merge_flux_rw$IDcrn)
site_name_sel <- c()
r_list <- c()
p_list <- c()

r_list_dtrd <- c()
p_list_dtrd <- c()

for(i in site_name) {
  tryCatch({
  df_sub <- subset(merge_flux_rw, IDcrn == i)
  
  r_corr_dtrd <- cor.test(df_sub$GPP_dtrd, df_sub$RWI_dtrd, method = "pearson")$estimate
  p_value_dtrd <- cor.test(df_sub$GPP_dtrd, df_sub$RWI_dtrd, method = "pearson")$p.value

  r_list_dtrd <- append(r_list_dtrd, r_corr_dtrd)
  p_list_dtrd <- append(p_list_dtrd, p_value_dtrd)
  
  site_name_sel <- append(site_name_sel, i)
  }, error = function(err) { })
}

#df_stat contains correlation for both detrended and non-detrended values
df_stat <- data.frame(site_name_sel, r_list_dtrd, p_list_dtrd)
df_stat$SITE_ID <- substr(df_stat$site_name_sel, 1,6)
df_stat <- df_stat %>% group_by(SITE_ID) %>% summarise(across(everything(), mean, na.rm = TRUE))

```


#only use lat lon without < -10 cWood, zero cWood, NA GPP and cWood
```{r}
model_name_list = c( "CLM5.0", "ISBA-CTRIP", "ORCHIDEE", "ORCHIDEEv3", "CLASSIC-N", "CLASSIC", "CABLE-POP", "LPX-Bern")

for(i in model_name_list) {
#Fluxes
gpp_df_S0 <- read.csv(paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", i, "/gpp_S0_31_site_weighted_yearly_mean.csv", sep = ""))
cWood_df_S0 <- read.csv(paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", i, "/cWood_S0_31_site_weighted_yearly_mean.csv", sep = ""))

gpp_df_S1 <- read.csv(paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", i, "/gpp_S1_31_site_weighted_yearly_mean.csv", sep = ""))
cWood_df_S1 <- read.csv(paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", i, "/cWood_S1_31_site_weighted_yearly_mean.csv", sep = ""))

gpp_df_S2 <- read.csv(paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", i, "/gpp_S2_31_site_weighted_yearly_mean.csv", sep = ""))
cWood_df_S2 <- read.csv(paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/TRENDYv10/31_site_weighted/", i, "/cWood_S2_31_site_weighted_yearly_mean.csv", sep = ""))

year_chosen = 1950
gpp_df_sub_S0 <- subset(gpp_df_S0, !is.na(gpp) & year > year_chosen)
cWood_df_sub_S0 <- subset(cWood_df_S0, !is.na(cWood) & year > year_chosen)

gpp_df_sub_S1 <- subset(gpp_df_S1, !is.na(gpp) & year > year_chosen)
cWood_df_sub_S1 <- subset(cWood_df_S1, !is.na(cWood) & year > year_chosen)

gpp_df_sub_S2 <- subset(gpp_df_S2, !is.na(gpp) & year > year_chosen)
cWood_df_sub_S2 <- subset(cWood_df_S2, !is.na(cWood) & year > year_chosen)

cWood_df_increment_S0 <- cWood_df_sub_S0 %>%
      arrange(lat, lon, year) %>% 
      group_by(lat, lon) %>%
      mutate(
            cWood_increment_S0 = (.data[["cWood"]] - lag(.data[["cWood"]]))/60/60/24/365,
            ) %>%
      ungroup()

cWood_df_increment_S1 <- cWood_df_sub_S1 %>%
      arrange(lat, lon, year) %>% 
      group_by(lat, lon) %>%
      mutate(
            cWood_increment_S1 = (.data[["cWood"]] - lag(.data[["cWood"]]))/60/60/24/365,
            ) %>%
      ungroup()

cWood_df_increment_S2 <- cWood_df_sub_S2 %>%
      arrange(lat, lon, year) %>% 
      group_by(lat, lon) %>%
      mutate(
            cWood_increment_S2 = (.data[["cWood"]] - lag(.data[["cWood"]]))/60/60/24/365,
            ) %>%
      ungroup()

merge_flux_S0 <- list(gpp_df_sub_S0, cWood_df_increment_S0) %>% 
                  reduce(full_join, by = c("year", "lat", "lon"))
merge_flux_S1 <- list(gpp_df_sub_S1, cWood_df_increment_S1) %>% 
                  reduce(full_join, by = c("year", "lat", "lon"))
merge_flux_S2 <- list(gpp_df_sub_S2, cWood_df_increment_S2) %>% 
                  reduce(full_join, by = c("year", "lat", "lon"))

merge_flux_sub_S0 <- subset(merge_flux_S0, !is.na(gpp) & !is.na(cWood_increment_S0) & cWood != 0)
merge_flux_sub_S1 <- subset(merge_flux_S1, !is.na(gpp) & !is.na(cWood_increment_S1) & cWood != 0)
merge_flux_sub_S2 <- subset(merge_flux_S2, !is.na(gpp) & !is.na(cWood_increment_S2) & cWood != 0)


merge_flux_sub_S0$cW_incre_gpp_ratio_S0 <- merge_flux_sub_S0$cWood_increment_S0/merge_flux_sub_S0$gpp*100
merge_flux_sub_S1$cW_incre_gpp_ratio_S1 <- merge_flux_sub_S1$cWood_increment_S1/merge_flux_sub_S1$gpp*100
merge_flux_sub_S2$cW_incre_gpp_ratio_S2 <- merge_flux_sub_S2$cWood_increment_S2/merge_flux_sub_S2$gpp*100

#extract the location where ratio is < -20
df_sub_S0 <- subset(merge_flux_sub_S0, cW_incre_gpp_ratio_S0 < -10)
df_sub_S1 <- subset(merge_flux_sub_S1, cW_incre_gpp_ratio_S1 < -10)
df_sub_S2 <- subset(merge_flux_sub_S2, cW_incre_gpp_ratio_S2 < -10)

problem_locations <- unique(
  rbind(
    df_sub_S0[, c("lat", "lon")],
    df_sub_S1[, c("lat", "lon")],
    df_sub_S2[, c("lat", "lon")]
  )
)

merge_flux_sub_S2 <- merge_flux_sub_S2 %>% anti_join(problem_locations, by = c("lat", "lon"))

merge_flux_sub_S2_sub <- merge_flux_sub_S2
merge_flux_sub_S2_sub$model <- i
lat_lon_list <- merge_flux_sub_S2_sub %>%
  distinct(lat, lon) 

#save the datasets with lat and longitude unique
path_save_coord = paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/datasets/",  i, "_", "coord_n_sample_TRENDYv10", ".csv", sep = "")
write.csv(lat_lon_list[, c("lat", "lon")], path_save_coord)
}


```


#reproduce cabon method with trendy dataset to calculate detrended and non detrended correlation
```{r}
model_name_list <- c("ISBA-CTRIP", "CABLE-POP", "CLM5.0", "LPX-Bern", "ORCHIDEE", "ORCHIDEEv3", "CLASSIC-N",  "CLASSIC")
scenario <- c("S0", "S1", "S2")
df_list_detrend <- list()
sample_num_list <- c()
for (i in 1: length(model_name_list)) {
  for (j in 1: length(scenario)) {
    path <- paste("/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/datasets/",  model_name_list[i], "_", "detrended_r_normal_delta_31_gpp_TRENDYv10_temporalMatch_not_scale", scenario[j], ".csv", sep = "")
    df <- read.csv(path)
    sample_num <- length(unique(df$SITE_ID))
    df$model <- model_name_list[i]
    df$scenario <- scenario[j]
    num <- 3*(i - 1) + j
    df_list_detrend[[num]]<- df
    sample_num_list <- append(sample_num_list, sample_num)
  }
}


```

```{r}
combined_df_detrend <- bind_rows(df_list_detrend)
plots <- list()

for(i in 1:length(model_name_list)) {

  data_sub <- subset(combined_df_detrend, model == model_name_list[i])
  
  df2 <- subset(df_stat, SITE_ID %in% unique(data_sub$SITE_ID)) %>%
    mutate(scenario = "obs")
  df2$r_detrend_delta <- df2$r_list_dtrd
  all_data <- bind_rows(data_sub, df2)
  all_data$model <- model_name_list[i]
  
  p_vals <- c(
    S0 = t.test(subset(all_data, scenario == "S0")$r_detrend_delta, subset(all_data, scenario == "obs")$r_list_dtrd)$p.value,
    S1 = t.test(subset(all_data, scenario == "S1")$r_detrend_delta, subset(all_data, scenario == "obs")$r_list_dtrd)$p.value,
    S2 = t.test(subset(all_data, scenario == "S2")$r_detrend_delta, subset(all_data, scenario == "obs")$r_list_dtrd)$p.value
  )
  
  signif_df <- data.frame(
  scenario = c("obs", "S0", "S1", "S2"),
  p_value = c(NA, p_vals),
  asterisk = c("", ifelse(p_vals < 0.05, "*", ""))
)
  
  signif <- signif_df %>%
    left_join(
      all_data %>%
        group_by(scenario) %>%
        summarise(y_pos = 0.05 + max(r_detrend_delta, na.rm = TRUE)),
      by = "scenario"
    )
  
  plots[[i]] <- 
   ggplot() + 
  geom_boxplot(
    data = all_data, 
    aes(x = scenario, y = r_detrend_delta), 
    size = 0.3, alpha = 0.5, outlier.shape = NA, color = "#3c5488", fill = "#3c5488"
  ) + geom_jitter(data = all_data, 
    aes(x = scenario, y = r_detrend_delta), 
    size = 0.3, alpha = 0.5, width = 0.15, color = "#3c5488", fill = "#3c5488") + 
  geom_text(
    data = signif,
    aes(x = scenario, y = y_pos, label = asterisk),
    color = "black",
    size = 5
  ) +
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 15),
    axis.text.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.title.x = element_text(size = 15),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  ) +
  ylim(-1, 1.2) +
  annotate(
    "text",
    x = 0.6,   
    y = -0.8,  
    label = paste(model_name_list[i], " (n = ", sample_num_list[i*3], ")", sep = ""),
    hjust = 0,
    vjust = 1,
    size = 3
  ) +
  ylab(expression(r[Delta * italic(cWood)* "," ~italic(GPP)]))
       
} 


#save all at once
save_path_boxplot = "/Users/ngocnguyen/Library/CloudStorage/Box-Box/Projects/biomass_paper/results/figures/all_delta_obs_r_gpp_TRENDYv10_temporalMatch_boxplot_31_non_scale.png"

png(save_path_boxplot,  width     = 8,
  height    = 4.25,
  units     = "in",
  res       = 1200,
  pointsize = 4)

#annotate for graph 1 and 5
for (i in 1:8) {
  if (i %% 4 != 1) {
    plots[[i]] <- plots[[i]] + theme(axis.title.y = element_blank(), axis.text.y = element_blank())
  }
  if(i <= 4) {
    plots[[i]] <- plots[[i]] + theme(axis.title.x = element_blank(), axis.text.x = element_blank())
  }
}

combined_plot <- (plots[[1]] + annotate("text",x = 0.7, y = 1.15, label = "a",size = 5) | 
                  plots[[2]]  + annotate("text",x = 0.7, y = 1.15, label = "b",size = 5) | 
                  plots[[3]]  + annotate("text",x = 0.7, y = 1.15, label = "c",size = 5) | 
                  plots[[4]]  + annotate("text",x = 0.7, y = 1.15, label = "d",size = 5)) /
                 (plots[[5]]  + annotate("text",x = 0.7, y = 1.15, label = "e",size = 5) | 
                  plots[[6]]  + annotate("text",x = 0.7, y = 1.15, label = "f",size = 5) | 
                  plots[[7]]  + annotate("text",x = 0.7, y = 1.15, label = "g",size = 5) | 
                  plots[[8]] + annotate("text",x = 0.7, y = 1.15, label = "h",size = 5) )
                
print(combined_plot)

dev.off()


```









